Binocular combination in the autonomic nervous system

### 2024-01-15 manuscript.Rmd
Updates to text and code

---
title: 'Chapter 4: Binocular combination in the Autonomic Nervous System'
author: |
 | Federico G. Segala$^1$, Joel T. Martin$^1$, Alex R. Wade$^1,2$ & Daniel H. Baker$^1,2$
output:
  bookdown::pdf_document2:
    fig_caption: yes
    toc: no
  html_document: default
bibliography: references.bib
link-citations: yes
linkcolor: black
csl: elife.csl
header-includes:
- \usepackage{caption}
---

@@

processdata <- 0  # this flag determines the amount of processing, with 4 levels:
# 0 - do no processing, generate the pdf using existing versions of all figures
# 1 - generate figures using the processed group data
# 2 - average data across participants, do bootstrapping
# 3 - download all raw data and analyse (requires 48GB of storage)
cleanupraw <- 0  # if downloading raw data, delete once processed to save space

@@

# cut off for data inclusion (Mahalanobis distance for complex data)
SDthresh <<- 3
SDthreshcond <- 1.85
nbootstraps <<- 1000
legaltriggers <- 1:55
targetelectrodes1 <- c('POz','POz')

# check which packages are installed, install the missing ones, and activate
packagelist <- c('knitr','remotes','tictoc','R.matlab','bookdown','grImport','png','tiff','pals','ez','gtools','signal','boot','quickpsy','rstan','rstatix','coda','parallel','utils','osfr','ggplot2','grid','gridExtra','cowplot','ggsignif') # list of CRAN packages

@@

# settings for sampler 
numSavedSteps <- 100000    # total number of MCMC steps
adaptSteps <- 1000  # Number of steps to "tune" the samplers
burnInSteps <- 2000 
thinSteps <- 10
nChains <- 16
ncores <- 8  # because of multi-threading, this is usually double the true number of physical cores, so we halve the number as Stan can only use physical cores
options(mc.cores=ncores)

@@

v4Interp <- function(df, xo, yo, rmax = .75, gridRes = 67) {
  ## Create a function to perform Matlab's v4 interpolation.
  ## Takes as input a data-frame with columns x, y, and z (x co-ordinates, y co-ordinates, and amplitude)
  ## and variables xo and yo, the co-ordinates which will be use to create a grid for interpolation
  xo <- matrix(rep(xo,length(yo)),nrow = length(xo),ncol = length(yo))
  yo <- t(matrix(rep(yo,length(xo)),nrow = length(yo),ncol = length(xo)))
  xy <- df$x + df$y*sqrt(as.complex(-1))
  d <- matrix(rep(xy,length(xy)),nrow = length(xy), ncol = length(xy))
  d <- abs(d - t(d))
  diag(d) <- 1
  g <- (d^2) * (log(d)-1)   # Green's function.
  diag(g) <- 0
  weights <- qr.solve(g,df$z)
  xy <- t(xy)
  outmat <- matrix(nrow = gridRes,ncol = gridRes)
  for (i in 1:gridRes){
    for (j in 1:gridRes) {
      test4 <- abs((xo[i,j] + sqrt(as.complex(-1))*yo[i,j]) - xy)
      g <- (test4^2) * (log(test4)-1)
      outmat[i,j] <- g %*% weights}}
  outDf <- data.frame(x = xo[,1],outmat)
  names(outDf)[1:length(yo[1,])+1] <- yo[1,]
  return(outDf)}

@@

if (processdata==2){
# if we're going to need the pre-processed files, extract from zip file
  if (!file.exists(paste0(datadir,'P101summary.RData'))){
   if (file.exists('resources/allprocessed.zip')){file.copy('resources/allprocessed.zip',localdir)}
   if (!file.exists('resources/allprocessed.zip')){ 
          osfnode <- '2qu6y'
          osfproject <- osf_retrieve_node(osfnode)
          osffiles <- osf_ls_files(osfproject,n_max=300)
          fid <- which(osffiles$name=='allprocessed.zip')
          osf_download(osffiles[fid,],localdir,progress=TRUE)
   }
          unzip(paste0(localdir,'allprocessed.zip'),exdir=localdir)
          d <- dir(paste0(localdir,'allprocessed'),full.names=TRUE)
          file.copy(d,datadir)
  }
}

@@

$^1$Department of Psychology, University of York, York, UK. YO10 5DD  
$^2$York Biomedical Research Institute, University of York, York, United Kingdom

@@

# Abstract

The size of the pupils changes in response to levels of ambient light and is regulated by the autonomic nervous system. The role of the retinal photoreceptors in determining pupil size has been investigated using silent substitution. They have been shown to be directly involved in controlling and maintaining the size of the pupils, with the cones and rods driving the initial constriction and the intrinsically photosensitive retinal ganglion cells maintaining the size over prolonged time periods. Here, we used silent substitution and pupillometry to investigate binocular combination while targeting the intrinsically-photoreceptive retinal ganglion cells, the L-M pathway and the S-(L+M) pathway. We found that the signals between the two pupils are combined in a non-linear manner when targeting all the different photoreceptor classes and that the non-linear combination was different between each class, as reflected by the different magnitudes of interocular suppression measured for each pathway.

The autonomic nervous system regulates many involuntary bodily processes, including the constriction and dilation of the pupils in response to light [@McDougal2015]. The anatomical pathway from the retina to the subcortical nuclei controlling the pupillary light response (PLR) is well established: it includes the Pretectal Olivary nucleus (PON), the Superior Cervical ganglion and the Edinger-Westphal nucleus, which project to the iris sphincter muscles that directly control the pupil size [@McDougal2008; @McDougal2015; @Wang2015]. Additionally, the retinal photoreceptors (cones, rods and the recently discovered melanopsin-containing intrinsically photosensitive retinal ganglion cells, ipRGCs) have been shown to be directly involved in controlling and maintaining the size of the pupils [@Barrionuevo2014; @Barrionuevo2018; @Dacey2005; @Murray2018; @Spitschan2014; @Spitschan2019; @Woelders2018]. The cones have been shown to drive the initial rapid constriction of the pupils [@Mathot2018], while the slower and longer activation of the ipRGCs shows their role in maintaining the constriction over a prolonged period of time and in regulating the post-illumination pupillary response [@Markwell2010; @McDougal2010].

The ipRGCs are a unique photoreceptor class that has been discovered recently [@Provencio2000]. They express the photopigment melanopsin, which is involved in the regulation of the circadian rhythm [@Provencio2000; @Panda2002; @Ruby2002], and form a major input to the PON [@Dacey2003]. The first direct evidence of the involvement of the ipRGCs in the PLR was shown in rats by @Lucas2003 who genetically deleted the melanopsin gene from rats, resulting in the loss of the intrinsic photosensitivity of the cells and a reduced pupil constriction. This same behaviour was later also observed in primates and humans [@Gamlin2007], where it was demonstrated that the PLR continues during light presentation even when cone and rod signalling is blocked. This also demonstrates the primary role of the ipRGCs in maintaining pupil constriction over a prolonged time. It has also been shown that the ipRGCs receive inputs from the other photoreceptors (cones and rods) and are activated by them [@Dacey2005].

To investigate how specific photoreceptor classes control the size of the pupils, a technique known as silent substitution is used. Silent substitution exploits the fact that each photoreceptor class has a distinct spectral tuning that overlaps with the others. Using a multiprimary system, in which the primaries have different spectra, it is possible to target one class of photoreceptors while maintaining the others at a constant activity level, effectively silencing them [@Shapiro1996; @Spitschan2018, this paper also offers a clear explanation of how to implement silent substitution]. Studies using silent substitution have shown that the pupils behave differently depending on which photoreceptor or pathway is being stimulated. For example, pupil responses have been shown to be out of phase when the S cones are stimulated, relative to when the ipRGCs or the L+M cones are stimulated [@Spitschan2014; @Cao2015]. Additionally, when L and M cones are stimulated individually, there appears to be a paradoxical pupil response: increments in L cone stimulation cause a constriction of the pupil, while increments in M cone stimulation lead to a dilation of the pupil [@Woelders2018; @Murray2018].

Interaction between the two eyes and the rules of binocular combination have been extensively studied. For pattern vision in the visual cortex, at low contrast, binocular summation occurs [binocular presentation confers higher sensitivity than monocular presentation, @Baker2018; @Campbell1965; @Legge1984], while, at high contrast, ocularity invariance is implemented [the response to monocularly- and binocularly-presented patterns is equalised, @Baker2007; @Ding2006; @Meese2006]. Additionally, evidence of a binocular component to the PLR is shown by the existence of a consensual response of the pupil [the stimulation of one eye will cause constriction of the other eye, @Wyatt1981] and the rules of binocular combination between the pupils have also been recently investigated [@Segala2023], finding that combination happens non-linearly with evidence of interocular suppression.

Most of the studies that use silent substitution and that investigate the PLR stimulate only one eye. If the stimulated eye is pharmacologically dilated to control pupil size, the response from the unstimulated eye (known as the consensual response) can be recorded [e.g. @McDougal2010; @Spitschan2014]. While this approach controls the total amount of light incident at the retina, it does not permit investigation of binocular combination for specific photoreceptor pathways. Seeing that it is possible to directly investigate binocular combination of signals between the pupils, we decided to investigate this by designing an experiment that allowed us to binocularly target the ipRGCs, the L-M pathway and the S-(L+M) pathway individually while recording binocular pupillometric measures. We chose a primary flicker frequency of 0.5 Hz, which has been shown to elicit pupil responses from all three pathways [@Spitschan2014]. The final results are interpreted using a hierarchical Bayesian computational model of binocular vision.

# Methods

Twenty-four participants were recruited for each of the four experiments for a total of ninety-six (68 females) adult participants, whose age ranged from 18 to 41. All participants had normal or corrected to normal binocular vision and normal colour vision, and gave written informed consent. Our procedures were approved by the Ethics Committee of the Department of Psychology at the University of York (identification number 184).

To present synchronised silent substitution contrast modulations independently to each eye, two light engines, each with 10 independently addressable LED colour channels (SpectraTuneLAB: LEDMOTIVE Technologies, LLC, Barcelona, Spain), were integrated into a customised viewing system. The light engines were operated via a Python interface to their REST API [@Martin2022], which supports synchronous launch and playback of spectral sequences prepared in advance and stored in JSON format. When preparing the spectral sequences, the age of participants was recorded to account for the yellowing of the lenses.

```{r allcons, fig.cap="Summary of the spectral power distributions and alpha-opic irradiances for the background and each experiment. Panel (a) shows a schematic of the binocular stimulation system for presenting spectrally tuned modulations independently to each eye. The VR headset was attached to a clamp stand that the experimenters could use to adjust the height and align the headset with the eyes of the participant. The participant's head was positioned on a chin rest to keep it in position throughout the entire experiment. Panels (b) and (c) show the spectral power distributions and the alpha-opic irradiances of the background spectra used for both eyes. Panels (d-k) show the power distributions and the alpha-opic irradiances relative to the background for the four experiments: luminance (d, e), melanopsin (f, g), L-M pathway (h, i) and S-(L+M) pathway (j, k). A positive difference power shows a positive modulation (white disc for the luminance experiment, magenta for the L-M pathway experiment and purple for the S-(L+M) experiment) and a negative difference power shows a negative modulation (black disc for the luminance experiment, cyan for the L-M pathway experiment and lime for the S-(L+M) experiment).", fig.align="center", out.width = '80%', echo=FALSE}

knitr::include_graphics(paste0(figdir,'allcons.pdf'))

The output from the light engines was directed through liquid light guides (LLG3-8H: Thorlabs Ltd, Cambridgeshire, UK) and diffused onto semi-opaque and highly diffusive white glass discs with a diameter of 50 mm for even illumination (34-473: Edmund Optics, York, UK). The light guide gaskets were butt-coupled to the light engine diffusers with threaded adapters (SM1A9, AD3LLG: Thorlabs Ltd, Cambridgeshire, UK) and the exiting ends of the light guides were mated with 51 mm depth optical cylinders (SM2L20: Thorlabs Ltd, Cambridgeshire, UK) via appropriately threaded adapters (AD3LLG, SM2A6: Thorlabs Ltd, Cambridgeshire, UK). The stimulus diffuser discs were retained at the front end of the optical cylinders approximately 51 mm from the light source, at which distance the output beam was sufficiently dispersed to afford even illumination of the diffuser when viewed from the front. To guarantee safe illumination levels, a circular neutral-density filter with the same diameter of the white glass discs (50 mm) and an optical density of 0.6 log units was placed in the optical path between the light source and the diffusers. A small circular piece of blackout material with a diameter of approximately 8 degrees (10 mm) was positioned centrally on the front of each diffuser disc to aid as a fusion lock, as a fixation point and to occlude the fovea.

The diffuser discs were positioned in the objective planes of the lenses of a modified VR headset (SHINECON SC-G01, Dongguan Shinecon Industrial Co. Ltd), which was used by the participants to view the stimuli. The stimuli were two discs of flickering light with a diameter of approximately 30 degrees, which were fused together into a cyclopean percept resembling a donut-shaped ring of light, similar to that used in other studies [e.g., @Barrionuevo2016; @Murray2018; @Spitschan2014; @Zele2018]. The VR headset modifications allowed for small adjustments to account for individual differences in interpupillary distance and focal length. The use of this set up allowed us to modulate the stimuli in three different ocular configurations, similar to the ones we used in our previous experiment [@Segala2023]: monocular, binocular and dichoptic. In the monocular configuration, the unstimulated eye still saw a non-flickering disc of mean luminance. A schematic of the stimulation system is shown in Figure \@ref(fig:allcons)a. Pupillometry data were collected using a binocular Pupil Core eye-tracker headset [Pupil Labs GmbH, Berlin, Germany, @Kassner2014] running at 120 Hz, and the signals were recorded with the Pupil Capture software.

Before the start of each experiment, participants adjusted the objective planes of the lenses with the help of the experimenter until the stimulus was in focus and they perceived the two pieces of blackout material as one fused piece.

```{r ExampleStims, fig.cap="Example of the different experimental conditions. Panel (a) shows one cycle of flicker in the achromatic luminance experiment. In this experiment, the stimulus flickers between black and white. At the beginning of the cycle (0s), the stimulus appears at half the intesity and appears grey to the viewer. At 0.5s, the stimulus reaches the maximum intensity of the flicker and appears white to the viewer. At 1s, the stimulus appears again at half maximum and once again appears to be grey to the viewer. At 1.5s, the stimulus reaches the minimum intensity of the flicker and appears black to the viewer. Panel (b) shows one cycle of the flickering in the L-M pathway experiment. In this experiment, the stimulus flickers between magenta (+L-M) and cyan (-L+M). At the beginning of the cycle (0s), the stimulus appears at zero cone contrast and appears to be grey to the viewer. At 0.5s, the stimulus reaches the maximum L-cone contrast and appears magenta to the viewer. At 1s, the stimulus appears again at zero cone contrast and switches colour, once again appearing to be grey to the viewer. At 1.5s, the stimulus reaches the minimum L-cone contrast and appears cyan to the viewer. Panel (c) shows one cycle of the flickering in the S-(L+M) pathway experiment. In this experiment, the stimulus flickers between purple (+S-(L+M)) and lime (-S+(L+M)). At the beginning of the cycle (0s), the stimulus appears at zero cone contrast and appears to be grey to the viewer. At 0.5s, the stimulus reaches the maximum S-cone contrast and appears purple to the viewer. At 1s, the stimulus appears again at zero cone contrast and switches colour, once again appearing to be grey to the viewer. At 1.5s, the stimulus reaches the minimum S-cone contrast and appears lime to the viewer.", fig.align="center", out.width = '95%', echo=FALSE}

knitr::include_graphics(paste0(figdir,'ExampleStims.pdf'))

We initially planned to examine pupil responses and EEG responses simultaneously to temporal modulations flickering at 2 Hz and 1.6 Hz while stimulating the periphery of the retina. While the experiment produced clear responses when presenting an achromatic light (see Figures \@ref(fig:appendixfig1) and \@ref(fig:appendixfig2) in Appendix 1), the responses were very noisy for the targeted photoreceptor class experiments (see Figures \@ref(fig:pupildataMel2Hz)-\@ref(fig:EEGdataS2Hz) in Appendix 2) for both the pupillometry and the EEG data. Therefore, we decided to lower the main frequency to 0.5 Hz as previous literature showed that it was slow enough to elicit a pupil response from all photoreceptor classes [@Spitschan2014] and we focussed on only recording pupillometry data as this frequency would be well below the minimum threshold to elicit EEG responses [@Norcia2015].

Pupil responses to binocular temporal contrast modulations were examined in a factorial design that combined six ocular conditions, two temporal frequencies (0.4 and 0.5 Hz) and five temporal contrast levels relative to the mean illuminance (6, 12, 24, 48 and 96%). This design, similar to what we used in our previous study [@Segala2023], was applied in four separate experiments, each with a different mode of photoreceptor stimulation. In the first three conditions, the discs flickered at 0.5 Hz, in either a monocular, binocular or dichoptic arrangement. In the dichoptic condition the non-target eye saw a flickering fixed contrast of 48%. In the remaining three conditions (the cross-frequency conditions) one eye’s disc flickered at 0.4 Hz, and the other eye’s disc flickered at 0.5 Hz. We also tested monocular responses at 0.4 Hz, as well as binocular (one eye sees each frequency at the target contrast) and dichoptic (target stimulus flickering at 0.5 Hz, mask contrast of 48% at 0.4 Hz in the other eye) arrangements. We counterbalanced presentation of the target stimulus across the left and right eyes.

For all experiments, sinusoidal contrast modulations were presented against the same background spectrum (matched between the eyes), which was used to achieve silent substitution in the three photoreceptor modulation experiments. The background spectra were defined by setting all channels to half maximum output for the brighter of the two devices (STLab 1, left eye) and then using the STLab 1/STLab 2 calibration ratio to find the equivalent settings for the companion device (STLab 2, right eye). The background spectrum illuminance was approximately 74 lux. The spectral power distributions and alpha-opic irradiances of the background spectra for both eyes are shown in Figures \@ref(fig:allcons)b-c.

### Experiment 1: luminance

The experiment was conducted in a windowless room, in which the only source of light was the modified VR headset. The participants sat as close as possible to the VR headset, leaving enough space for the eye-tracker to record the eyes. The experiment was carried out in a single session of 45 minutes, divided into three blocks of 15 minutes each. In each block, there were a total of 60 trials lasting 15 seconds each (12s of stimulus presentation, followed by 3s of interstimulus interval). The participants were given no task other than look at the black fixation dot while trying to minimise their blinking during the presentation period.

In this experiment, the participants saw a disc of achromatic flickering light (average illuminance of approximately 74 lux), flickering between black and white, at the different contrast levels defined earlier. An example of one cycle of flickering for the luminance experiment is shown in Figure \@ref(fig:ExampleStims)a. The spectral power distributions and alpha-opic irradiances relative to the background for the luminance experiment are shown in Figures \@ref(fig:allcons)d-e.

### Experiment 2: melanopsin

The experiment used the same equipment set-up and the same experimental conditions as Experiment 5. The procedure was also the same. However, at the beginning of each block, participants habituated for 2 minutes to the background that would be present throughout the entire experiment. During this adaptation period, participants were asked to fixate on the same black fixation dot positioned in the middle of the stimulus and were told that they could blink normally. At the end of the adaptation period, the experiment started immediately and the participants followed the same instructions given in Experiment 1.

In this experiment, the participants experienced contrast modulations that targeted the melanopsin-containing intrinsically photosensitive Retinal Ganglion cells, while ignoring rods and nominally silencing the cones. Michelson contrast levels represented a percentage of the maximum melanopic contrast available, which was predetermined to be approximately 22 \%. Under perfect silent substitution conditions, these modulations do not entail any changes in brightness or chromaticity: it is therefore very difficult to include an example to visualise the stimulus. However, silent substitution is never perfect, so participants reported small fluctuations in reddishness, greenishness and brightness due to open-field cone contrast artefacts. These fluctuations were especially present at higher contrasts. This can be observed in the alpha-opic irradiances relative to the background for the melanopsin experiment shown in Figure \@ref(fig:allcons)g, where a small activation of the S, L and M cones is observable. The spectral power distributions relative to the background for the melanopsin experiment is also shown in Figure \@ref(fig:allcons)f.

### Experiment 3: L-M postreceptoral ‘parvocellular’ pathway

The experiment used the same equipment set-up and the same experimental conditions as Experiment 5. The procedure was the same as Experiment 6, including the presence of a 2-minute adaptation period at the beginning of each block. Additionally, before the start of the experiment, participants completed a luminance nulling perceptual calibration procedure in L-M cone space on an Iiyama $\textrm{VisionMaster}^{\textrm{TM}}$ Pro 510 display (800 x 600 pixels, 60 Hz refresh rate). During the task, participants were presented with a disc flickering within the L-M cone space (between magenta and cyan). Using a trackball, participants adjusted the angle in cone space to find their subjective isoluminant point, which resulted in changing the flickering intensity of the stimulus until the amplitude of the flicker appeared to be minimised. The result was used to modify the requested contrasts during stimulus preparation so as to account for individual differences affecting perceived illuminance, principally the L:M cone ratio [@Carroll2002; @Hofer2005].

In this experiment, the participants experienced contrast modulations that targeted the L-M postreceptoral ‘parvocellular’ pathway while ignoring rods and nominally silencing S-cones and melanopsin. Michelson contrast levels represented a percentage of the maximum L-M contrast available for the background spectrum, which was determined in advance to be approximately 10 \%. As shown in Figure \@ref(fig:ExampleStims)b, participants experienced an isoluminant chromatic modulation between magenta (+L-M) and cyan (-L+M). The spectral power distributions and alpha-opic irradiances relative to the background for the L-M pathway experiment are shown in Figures \@ref(fig:allcons)h-i.

### Experiment 4: S-(L+M) ‘koniocelluar’ pathway

The experiment used the same equipment set-up and the same experimental conditions as Experiment 5. The procedure was the same as Experiment 6, including the presence of a 2-minute adaptation period at the beginning of each block.

In this experiment, the participants experienced contrast modulations that targeted the S-(L+M) ‘koniocelluar’ pathway while ignoring rods and nominally silencing L/M cones and melanopsin. Michelson contrast levels represented a percentage of the maximum S-cone contrast available for the background spectrum, which was predetermined to be approximately 45 \%. As shown in Figure \@ref(fig:ExampleStims)c, participants experienced an isoluminant chromatic modulation between purple (+S-(L+M)) and lime (-S+(L+M)). The spectral power distributions and alpha-opic irradiances relative to the background for the S-(L+M) pathway experiment are shown in Figures \@ref(fig:allcons)j-k.

The pupillometry data were analysed using the same method we used in our previous study [@Segala2023]. The data were converted from mp4 videos to a csv text file using the Pupil Player software [@Kassner2014], which estimated pupil diameter for each eye on each frame using a 3D model of the eyeball. The individual data were then loaded into R for analysis, where a ten-second waveform for each trial in each eye was extracted (excluding the first two seconds after stimulus onset). We interpolated across any dropped or missing frames to ensure regular and continuous sampling over time. The Fourier transform was calculated for each waveform, and all repetitions of each condition were pooled across eye and then averaged. Finally, data were averaged across all participants to obtain the group results. We used coherent averaging and at each stage we excluded data points with a Mahalanobis distance exceeding $D$ = `r SDthresh` from the complex-valued mean. The consensual response was also analysed (see Appendix 3 for more details and the results).

To describe our data, we chose the same model described in our previous study [@Segala2023]. The model has the same general form as the first stage of the contrast gain control model proposed by @Meese2006 and omits the second stage. For the previous model that we used [@Segala2023], the exponent of the numerator had a fixed value of 2. Here, we allow this parameter (called _p_) to be free, in order to permit different shapes of contrast response function, e.g. saturating or super-saturating. The responses of the left eye and right eye channels are as follows:

\noindent where _n_ is a noise parameter, and $R_{max}$ scales the overall response amplitude.

We performed model fitting on the normalised amplitudes to account for amplitude differences between the experiments.

## Experiment 1

```{r luminance individual, include=FALSE, results='hide'}

# (download and) analyse individual participant data for the luminance experiment
if (processdata > 2){

  for(participant in 1:24){

    # only download and process data if this has not been done already for this participant
    if (!file.exists(paste0(datadir,'P',600+participant,'summary.RData'))){

      # if there is no directory for this participant's data
      if (!file.exists(paste0(rawdir,'P',600+participant))){
        dir.create(paste0(rawdir,'P',600+participant))

       d <- dir(paste0(rawdir,'P',600+participant))

       if (length(d)==0){

         # if the tar file of this participant's data doesn't exist
         if (!file.exists(paste0(rawdir,'P',600+participant,'.zip'))){
           # download it from OSF
           if (!exists('osffiles')){
           osfnode <- '6u7db'
           osfproject <- osf_retrieve_node(osfnode)
          osffiles <- osf_ls_files(osfproject,n_max=300)}
          fid <- which(osffiles$name==paste0('P',600+participant,'.zip'))
          osf_download(osffiles[fid,],rawdir,progress=TRUE)
         }

         # then unzip the tar file
         unzip(paste0(rawdir,'P',600+participant,'.zip'),exdir=paste0(rawdir))

         # and delete the tar file to save storage space
         file.remove(paste0(rawdir,'P',600+participant,'.zip'))
       }
      }

      d <- dir(paste0(rawdir,'P',600+participant),full.names=TRUE)

    tic()

@@

    repcounterP <- (1:60)*0
    afiles <- dir(path=paste0(rawdir,'P',600+participant),pattern='*annotations.csv',full.names=TRUE)

    for (block in 1:length(afiles)){
        tic()
        adata <- read.csv(paste0(rawdir,'P',600+participant,'/',paste0('P',600+participant,'_S',block),'_annotations.csv'))

@@

        pdata <- read.csv(paste(rawdir,'P',600+participant,'/',paste0('P',600+participant,'_S',block),'_pupil_positions.csv',sep=''))
        pdata2 <- pdata[,c(1,3,4,14)]
        pdata2 <- pdata2[which(pdata2[,3]>0.2),]

@@

    meanwavesP <- apply(pupilwaveforms[,,c(15,20),],4,mean,na.rm=TRUE)
    meanspectraP <- apply(pupilspectra[,,c(15,20),],4,mean,na.rm=TRUE)
    save(file=paste(datadir,'P',600+participant,'_summaryLum.RData',sep=''),list=c('cleanmeansP','cleanmasksP','cleanmeansP2','cleanmasksP2','meanspectraP','meanwavesP'))

    if (cleanupraw==1){file.remove(d)} # delete raw data files to save space
  }

  }}
```

```{r luminance group, include=FALSE, results='hide'}

# load in individual participant data and average, save group data
if (processdata > 1){
  for (participant in 1:24){
         # if the summary file of this participant's data doesn't exist
         if (!file.exists(paste0(datadir,'P',600+participant,'summaryLum.RData'))){
           # download it from OSF
           if (!exists('osffilesP')){
           osfnode <- '9ebtd'
           osfproject <- osf_retrieve_node(osfnode)
          osffilesP <- osf_ls_files(osfproject,n_max=300)}
          fid <- which(osffilesP$name==paste0('P',600+participant,'summaryLum.RData'))
          osf_download(osffilesP[fid,],datadir,progress=TRUE)
         }
}

  d <- dir(datadir,pattern='summaryLum.RData', full.names = TRUE)

@@

  tempspectraP <- tempspectraP[-c(2,4,13,20,26),]

@@

      temp <- tempmeansP[,cond,level]
      cartdata <- data.frame(Re(temp),Im(temp))
      if (sum(cartdata)!=0){
        D <- sqrt(stats::mahalanobis(cartdata, colMeans(cartdata), cov(cartdata)))
        i <- which(D<SDthresh)
        mbs <- amperrors(temp[i],method='boot',quantiles=68,nresamples=nbootstraps)
        cleanmeansP[cond,level] <- mean(temp[i])
        cleanmeansPCI[1,cond,level] <- mbs$lowerCI
        cleanmeansPCI[2,cond,level] <- mbs$upperCI
      }

@@

      temp <- tempmasksP2[,cond,level]
      cartdata <- data.frame(Re(temp),Im(temp))
      if (sum(cartdata)!=0){
        D <- sqrt(stats::mahalanobis(cartdata, colMeans(cartdata), cov(cartdata)))
        i <- which(D<SDthresh)
        mbs <- amperrors(temp[i],method='boot',quantiles=68,nresamples=nbootstraps)
        cleanmasksP2[cond,level] <- mean(temp[i])
        cleanmasksP2CI[1,cond,level] <- mbs$lowerCI
        cleanmasksP2CI[2,cond,level] <- mbs$upperCI
      }
    }
  }

@@

```{r luminance plot, include=FALSE, results='hide'}

# generate plot showing the group pupillometry data for the luminance experiment

         if (!file.exists(paste0(datadir,'AveragedataLum.RData'))){
           # download it from OSF
           if (!exists('osffilesP')){
           osfnode <- '9ebtd'
           osfproject <- osf_retrieve_node(osfnode)
          osffilesP <- osf_ls_files(osfproject,n_max=300)}
          fid <- which(osffilesP$name=='AveragedataLum.RData')
          osf_download(osffilesP[fid,],datadir,progress=TRUE)
         }

>>>>

---
title: 'Binocular combination in the autonomic nervous system'
author: |
 | Federico G. Segala$^1$, Joel T. Martin$^1$, Annie Y. Morsi$^1$, Alex R. Wade$^{1,2}$ & Daniel H. Baker$^{1,2}$
output:
  bookdown::pdf_document2:
    fig_caption: yes
    toc: no
    keep_tex: true
  html_document: default
bibliography: references.bib
link-citations: yes
linkcolor: black
csl: elife.csl
header-includes:
- \usepackage{caption}
---

@@

processdata <- 1  # this flag determines the amount of processing, with 4 levels:
# 0 - do no processing, generate the pdf using existing versions of all figures
# 1 - generate figures using the processed group data
# 2 - average data across participants, do bootstrapping
# 3 - download all raw data and analyse (requires 56GB of storage)

@@

# cut off for data inclusion (Mahalanobis distance for complex data)
SDthresh <<- 3
nbootstraps <<- 1000

# check which packages are installed, install the missing ones, and activate
packagelist <- c('knitr','remotes','tictoc','R.matlab','bookdown','grImport','png','tiff','pals','ez','gtools','signal','boot','quickpsy','rstan','bayestestR','rstatix','coda','parallel','utils','osfr','ggplot2','grid','gridExtra','cowplot','ggsignif','osfr') # list of CRAN packages

@@

# settings for sampler 
numSavedSteps <- 10000    # total number of MCMC steps
adaptSteps <- 1000  # Number of steps to "tune" the samplers
burnInSteps <- 2000 
thinSteps <- 10
nChains <- 16
ncores <- 8  # because of multi-threading, this is usually double the true number of physical cores, so we halve the number as Stan can only use physical cores
options(mc.cores=ncores)

@@

@@

# these are approximate values for isolumiant red, green, blue and yellow
stimcolours <- c(rgb(0.02,0.02,0.02),rgb(0.98,0.98,0.98),rgb(0,0.61,0.5),rgb(1,0.39,0.5),rgb(0.35,0.63,0),rgb(0.65,0.37,1),'grey','cornflowerblue')

if (processdata>2){
  # check if raw data are present locally, and download if not
  nodelist <- c('6u7db','xkgd9','pexyk','k9mdq')  # list of OSF nodes containing raw data
  for (expt in 1:4){
    osfproject <- osf_retrieve_node(nodelist[expt])
    osffiles <- osf_ls_files(osfproject,n_max=300)
    for (f in 1:nrow(osffiles)){
      fname <- osffiles$name[f]
      foldername <- strsplit(fname,'.',fixed=TRUE)[[1]][1]
      if (!dir.exists(paste0(rawdir,foldername))){
        osf_download(osffiles[f,],localdir,progress=TRUE)
        dir.create(paste0(rawdir,foldername))
        unzip(paste0(localdir,fname),exdir=paste0(rawdir,foldername))
        file.remove(paste0(localdir,fname))
      }
    }
  }
}

# if (processdata==2){
# # if we're going to need the pre-processed files, extract from zip file
#   if (!file.exists(paste0(datadir,'P101summary.RData'))){
#    if (file.exists('resources/allprocessed.zip')){file.copy('resources/allprocessed.zip',localdir)}
#    if (!file.exists('resources/allprocessed.zip')){ 
#           osfnode <- '2qu6y'
#           osfproject <- osf_retrieve_node(osfnode)
#           osffiles <- osf_ls_files(osfproject,n_max=300)
#           fid <- which(osffiles$name=='allprocessed.zip')
#           osf_download(osffiles[fid,],localdir,progress=TRUE)
#    }
#           unzip(paste0(localdir,'allprocessed.zip'),exdir=localdir)
#           d <- dir(paste0(localdir,'allprocessed'),full.names=TRUE)
#           file.copy(d,datadir)
#   }
# }

@@

$^1$Department of Psychology, University of York, Heslington, York, United Kingdom.

$^2$York Biomedical Research Institute, University of York, Heslington, York, United Kingdom.

@@

```{r individualanalyses, include=FALSE, results='hide'}

# analyse individual participant data for all conditions

if (processdata > 2){

  allsubjects <- dir(rawdir)
  for (p in 1:length(allsubjects)){
    participant <- allsubjects[p]

@@

    repcounterP <- (1:60)*0
    annfiles <- dir(paste0(rawdir,participant),pattern='*annotations.csv',full.names=TRUE)
    pupilfiles <- dir(paste0(rawdir,participant),pattern='*pupil_positions.csv',full.names=TRUE)

    for (block in 1:length(annfiles)){

        adata <- read.csv(annfiles[block])

@@

        pdata <- read.csv(pupilfiles[block])
        pdata2 <- pdata[,c(1,3,4,14)]
        pdata2 <- pdata2[which(pdata2[,3]>0.2),]

@@

    meanwavesP <- apply(pupilwaveforms[,,c(15,20),],4,mean,na.rm=TRUE)
    meanspectraP <- apply(pupilspectra[,,c(15,20),],4,mean,na.rm=TRUE)
    save(file=paste(datadir,participant,'_summary.RData',sep=''),list=c('cleanmeansP','cleanmasksP','cleanmeansP2','cleanmasksP2','meanspectraP','meanwavesP'))
  }

```{r groupanalyses, include=FALSE, results='hide'}

# files from P301-P324 are for S-cone stimulation
# files from P501-P524 are for melanopsin stimulation
# files from P601-P624 are for luminance stimulation
# files from P701-P724 are for (L-M) cone stimulation

# load in individual participant data and average, save group data
if (processdata > 1){

  prefixlist <- c('P6','P7','P3','P5')
  exptnames <- c('AveragedataLum.RData','AveragedataLM.RData','AveragedataS.RData','AveragedataMel.RData')

  includedvalues <- array(0,dim=c(4,24,6,5))
  includedvalues2 <- array(0,dim=c(4,24,6,5))

  for (expt in 1:4){
  d <- dir(datadir,pattern=prefixlist[expt], full.names = TRUE)

@@

  # tempspectraP <- tempspectraP[-c(2,4,13,20,26),]

@@

      temp <- tempmeansP[,cond,level]
      cartdata <- data.frame(Re(temp),Im(temp))
      if (sum(cartdata)!=0){
        D <- sqrt(stats::mahalanobis(cartdata, colMeans(cartdata), cov(cartdata)))
        i <- which(D<SDthresh)
        includedvalues[expt,i,cond,level] <- 1
        mbs <- amperrors(temp[i],method='boot',quantiles=68,nresamples=nbootstraps)
        cleanmeansP[cond,level] <- mean(temp[i])
        cleanmeansPCI[1,cond,level] <- mbs$lowerCI
        cleanmeansPCI[2,cond,level] <- mbs$upperCI
      }

@@

      temp <- tempmasksP2[,cond,level]
      cartdata <- data.frame(Re(temp),Im(temp))
      if (sum(cartdata)!=0){
        D <- sqrt(stats::mahalanobis(cartdata, colMeans(cartdata), cov(cartdata)))
        i <- which(D<SDthresh)
        includedvalues2[expt,i,cond,level] <- 1
        mbs <- amperrors(temp[i],method='boot',quantiles=68,nresamples=nbootstraps)
        cleanmasksP2[cond,level] <- mean(temp[i])
        cleanmasksP2CI[1,cond,level] <- mbs$lowerCI
        cleanmasksP2CI[2,cond,level] <- mbs$upperCI
      }
    }
  }

@@
